Comparing clustering parameters with rOpenScPCA

### 2025-01-13 analyses/hello-clusters/02_compare-clustering-parameters.Rmd
flesh out section to vary one parameter

* Calculate clusters across several different parameterizations
* Calculate QC metrics on across parameterizations

@@

* Calculate clusters from a single parameterization
* Calculate QC metrics on a single set of clusters

@@

suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(Seurat)
  library(dplyr)
  library(ggplot2)
})

@@

This function takes a PCA matrix with rownames representing unique cell ids (e.g., barcodes) as its primary argument, with additional arguments for cluster parameters. 
This function wraps the `rOpenScPCA::calculate_clusters()` function but allows you to provide a vector of parameter values to perform clustering across, as listed below.
Clusters will be calculated for all combinations of parameters values (where applicable); default values that the function will use for any unspecified parameter values are shown in parentheses

@@

```{r sweep clusters}
# Define nn parameter values of interest
nn <- seq(10, 30, 10)

# Calculate clusters varying nn, but leaving other parameters at their default values
cluster_results_list <- rOpenScPCA::sweep_clusters(
  pca_matrix,
  nn = nn
)
```

@@

@@

Generally speaking, `purrr::map()` can be used to iterate over this list to analyze or visualize each clustering result on its own.
Complementary to this, we can also combine this list into a single data frame to jointly analyze or visualize all clustering results together:

```{r bind cluster_results_list rows}
cluster_results_df <- dplyr::bind_rows(cluster_results_list)
head(cluster_results_df)
```

@@

This section will use `purrr::map()` to iterate over each df to calculate silhouette, purity, and stability. 
We'll also plot results.

>>>>

* Calculate several versions of clustering results across several different parameterizations
* Calculate QC metrics on across clustering results

@@

* Calculate clusters from a single parameterization
* Calculate QC metrics on a single set of clusters, as well as explanations of the metrics themselves

@@

suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(ggplot2)
  library(patchwork)
})

@@

This function takes a PCA matrix with row names representing unique cell ids (e.g., barcodes) as its primary argument, with additional arguments for cluster parameters. 
This function wraps the `rOpenScPCA::calculate_clusters()` function but allows you to provide a vector of parameter values to perform clustering across, as listed below.
Clusters will be calculated for all combinations of parameters values (where applicable); default values that the function will use for any unspecified parameter values are shown in parentheses

@@

```{r sweep clusters}
# Define nn parameter values of interest
nn_values <- seq(10, 30, 10)

# Calculate clusters varying nn, but leaving other parameters at their default values
cluster_results_list <- rOpenScPCA::sweep_clusters(
  pca_matrix,
  nn = nn_values
)
```

@@

It can be helpful (although it is not strictly necessary to keep track) to name this list by the varied `nn` parameter:

```{r set list names}
names(cluster_results_list) <- glue::glue("nn_{nn_values}")
```

@@

Generally speaking, `purrr::map()` can be used to iterate over this list to visualize or analyze each clustering result on its own; we'll use this approach in the following sections.

## Visualizing clustering results

When comparing clustering results, it's important to first visualize the different clusterings to build context for interpreting QC metrics.

As one example of why this is important, we generally expect that more robust clusters will have higher values for metrics like silhouette width and neighborhood purity.
However, we also expect that having fewer clusters in the first place will also lead to higher metrics, regardless of cluster quality: When there are fewer clusters, it is more likely that clusters overlap less with one another just because there aren't many clusters in the first place.
This means that, when interpreting cluster quality metrics, you should be careful to take more context about the data into consideration and not only rely on the metric values.

We'll therefore visualize these results as UMAPs by iterating over `cluster_results_list` and combining plots with [`patchwork::wrap_plots()`](https://patchwork.data-imaginist.com/reference/wrap_plots.html).
We'll specifically use [`purrr::imap()`](https://purrr.tidyverse.org/reference/imap.html) to iterate so that we can assign this list's names as plot titles.

For this, we'll begin by extracting a table of UMAP coordinates from our SCE object.

```{r create umap_df}
umap_df <- reducedDim(sce, "UMAP") |>
  as.data.frame()
```

Next, we'll iterate over `cluster_results_list` to plot the UMAPs.

```{r plot all umaps, fig.width = 12}
umap_plots <- cluster_results_list |>
  purrr::imap(
    \(cluster_df, clustering_name) {
      # Add a column with cluster assignments to umap_df
      umap_df_plot <- umap_df |>
        dplyr::mutate(cluster = cluster_df$cluster)

      # Plot the UMAP, colored by the new cluster variable
      ggplot(umap_df_plot, aes(x = UMAP1, y = UMAP2, color = cluster)) +
        geom_point(alpha = 0.6) +
        labs(title = glue::glue("nearest neighbors: {clustering_name}")) +
        # We'll add a couple UMAP plot settings here, including equal axes and
        # turning off the axis ticks and text since UMAP coordinates are not meaningful
        coord_equal() +
        theme(
          axis.ticks = element_blank(),
          axis.text = element_blank()
        )
    }
  )

# Print the plots with patchwork::wrap_plots()
patchwork::wrap_plots(umap_plots)
```

These plots show that the number of clusters decreases as the nearest neighbors parameter increases, with between 9-13 clusters.

@@

This section will use `purrr::map()` to iterate over each clustering result data frame to calculate silhouette width, neighborhood purity, and stability, and then visualize results.
The goal of this code is to identify whether one clustering parameterization produces more reliable clusters.

### Silhouette width and neighborhood purity

Both silhouette width and neighborhood purity are cell-level quantities, so we can calculate them together in the same call to `purrr::map()`.
Below, we'll iterate over each data frame in `cluster_results_list` to calculate these quantities.

```{r calculate cell level metrics}
cell_metric_list <- cluster_results_list |>
  purrr::map(
    \(cluster_df) {
      # calculate silhouette width
      silhouette_df <- rOpenScPCA::calculate_silhouette(pca_matrix, cluster_df)

      # calculate neighbhorhood purity
      purity_df <- rOpenScPCA::calculate_purity(pca_matrix, cluster_df)

      # Combine into a single data frame
      dplyr::left_join(silhouette_df, purity_df)
    }
  )

# View the first six rows of each clustering result's cell-level QC metrics
purrr::map(cell_metric_list, head)
```

To visualize these results, we can combine all data frames in this list into a single overall data frame, where the existing `nn` column will distinguish among conditions.

```{r combine cell metrics list}
cell_metrics_df <- dplyr::bind_rows(cell_metric_list)
```

We can visualize silhouette width and neighborhood purity each with boxplots, for example, and use the [`patchwork`](https://patchwork.data-imaginist.com/) package to print them together:

```{r}
# Plot silhouette width
silhouette_plot <- ggplot(cell_metrics_df) +
  aes(x = as.factor(nn), y = silhouette_width, fill = as.factor(nn)) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Pastel2") +
  labs(
    x = "Number of nearest neighbors",
    y = "Silhouette width"
  )

# Plot neighborhood purity width
purity_plot <- ggplot(cell_metrics_df) +
  aes(x = as.factor(nn), y = purity, fill = as.factor(nn)) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Pastel2") +
  labs(
    x = "Number of nearest neighbors",
    y = "Neighborhood purity"
  )

# Add together using the patchwork library, without a legend
silhouette_plot + purity_plot & theme(legend.position = "none")
```
While there does not appear to be a salient difference among silhouette width distributions, it does appear that purity is higher with a higher nearest neighbors parameter.

### Stability

Next, we'll calculate stability on the clusters using `rOpenScPCA::calculate_stability()`, specifying the same parameter used for the original cluster calculation at each iteration.

```{r calculate stability}
stability_list <- cluster_results_list |>
  purrr::map(
    \(cluster_df) {
      nn <- unique(cluster_df$nn)

      # calculate stability, passing in the parameter value used for this iteration
      rOpenScPCA::calculate_stability(pca_matrix, cluster_df, nn = nn)
    }
  )
```

We'll again combine the output of `stability_list` into a single data frame and plot `ari` values across `nn` parameterizations.

```{r combine plot stability}
stability_df <- dplyr::bind_rows(stability_list)

ggplot(stability_df) +
  aes(x = as.factor(nn), y = ari, fill = as.factor(nn)) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Pastel2") +
  labs(
    x = "Number of nearest neighbors",
    y = "Adjusted Rand Index"
  ) +
  theme(legend.position = "none")
```

Here, we see that a nearest neighbors value of 20 or 30 leads to more stable clustering results compared to 10.
